---
title: "Vignette 2: Analyze and Visualize Results"
author: "Rebecca C. Hendrickson, Ronald G. Thomas"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Vignette 2: Analyze and Visualize Results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 7,
  eval = TRUE
)
```

## Overview

This vignette demonstrates how to analyze and visualize simulation results
from pmsimstats2025. The workflow covers:

1. Loading and exploring simulation results
2. Computing summary statistics (power, effect sizes)
3. Creating power heatmaps
4. Comparing designs with and without carryover modeling
5. Generating publication-quality figures

## Setup

```{r load-packages, message=FALSE}
rm(list = ls())
suppressPackageStartupMessages({
  library(tidyverse)
  library(ggplot2)
  library(patchwork)
  library(viridis)
})
```

## Part 1: Load Simulation Results

```{r load-results}
# ============================================================================
# LOAD RESULTS
# ============================================================================

# Load vignette results (generated by Vignette 1)
base::load("../analysis/output/vignette_simulation_results.RData")
system("ls ../")
cat("Loaded objects:\n")
cat(sprintf("  - results: %d rows\n", nrow(results)))
cat(sprintf("  - summary_results: %d rows\n", nrow(summary_results)))
cat(sprintf("  - param_grid: %d conditions\n", nrow(param_grid)))
```

### Explore Data Structure

```{r explore-data}
# ============================================================================
# EXPLORE DATA STRUCTURE
# ============================================================================

cat("\n")
cat("=" %+% strrep("=", 59) %+% "\n")
cat("DATA STRUCTURE\n")
cat("=" %+% strrep("=", 59) %+% "\n\n")

cat("Results columns:\n")
cat(paste("  -", names(results), collapse = "\n"), "\n\n")

cat("Designs in dataset:\n")
cat(paste("  -", unique(results$design), collapse = "\n"), "\n\n")

cat("Biomarker moderation values:\n")
cat(paste("  -", unique(results$biomarker_moderation), collapse = ", "), "\n\n")

cat("Carryover values:\n")
cat(paste("  -", unique(results$carryover), collapse = ", "), "\n")
```

## Part 2: Compute Summary Statistics

```{r summary-stats}
# ============================================================================
# SUMMARY STATISTICS
# ============================================================================

cat("\n")
cat("=" %+% strrep("=", 59) %+% "\n")
cat("POWER ANALYSIS SUMMARY\n")
cat("=" %+% strrep("=", 59) %+% "\n\n")

# Power by design
power_by_design <- summary_results %>%
  group_by(design) %>%
  summarize(
    n_conditions = n(),
    mean_power = mean(power, na.rm = TRUE),
    sd_power = sd(power, na.rm = TRUE),
    min_power = min(power, na.rm = TRUE),
    max_power = max(power, na.rm = TRUE),
    .groups = "drop"
  )

cat("Power by Design:\n")
print(power_by_design)

# Power by biomarker moderation
cat("\n\nPower by Biomarker Moderation:\n")
power_by_bm <- summary_results %>%
  group_by(biomarker_moderation) %>%
  summarize(
    n_conditions = n(),
    mean_power = mean(power, na.rm = TRUE),
    .groups = "drop"
  )
print(power_by_bm)

# Power by carryover
cat("\n\nPower by Carryover:\n")
power_by_co <- summary_results %>%
  group_by(carryover) %>%
  summarize(
    n_conditions = n(),
    mean_power = mean(power, na.rm = TRUE),
    .groups = "drop"
  )
print(power_by_co)
```

### Type I Error Analysis

```{r type1-error}
# ============================================================================
# TYPE I ERROR ANALYSIS
# ============================================================================

cat("\n")
cat("=" %+% strrep("=", 59) %+% "\n")
cat("TYPE I ERROR RATES\n")
cat("=" %+% strrep("=", 59) %+% "\n\n")

cat("(Conditions where biomarker_moderation = 0)\n\n")

type1_error <- summary_results %>%
  filter(biomarker_moderation == 0) %>%
  select(design, carryover, power) %>%
  mutate(
    type1_rate = sprintf("%.1f%%", power * 100),
    nominal = ifelse(power <= 0.05, "OK", "ELEVATED")
  )

print(type1_error)
```

## Part 3: Create Power Heatmaps

### Single Design Heatmap

```{r single-heatmap}
# ============================================================================
# SINGLE DESIGN HEATMAP
# ============================================================================

plot_power_heatmap_single <- function(data, design_name) {
  design_labels <- c(
    "ol_bdc" = "OL+BDC\n(Design 2)",
    "hybrid" = "Hybrid/N-of-1\n(Design 4)"
  )

  plot_data <- data %>%
    filter(design == design_name) %>%
    mutate(
      carryover = factor(carryover),
      biomarker_moderation = factor(
        biomarker_moderation,
        levels = rev(sort(unique(biomarker_moderation)))
      )
    )

  p <- ggplot(
    plot_data,
    aes(x = carryover, y = biomarker_moderation, fill = power)
  ) +
    geom_tile(color = "white", linewidth = 0.5) +
    geom_text(
      aes(label = sprintf("%.0f%%", power * 100)),
      color = "black",
      size = 4,
      fontface = "bold"
    ) +
    scale_fill_gradient2(
      name = "Power",
      low = "#d73027",
      mid = "#fee08b",
      high = "#1a9850",
      midpoint = 0.5,
      limits = c(0, 1),
      labels = scales::percent
    ) +
    scale_x_discrete(expand = c(0, 0)) +
    scale_y_discrete(expand = c(0, 0)) +
    labs(
      title = design_labels[design_name],
      x = "Carryover (proportion retained)",
      y = "Biomarker Moderation\n(effect size)"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      panel.grid = element_blank(),
      axis.ticks = element_blank(),
      plot.title = element_text(hjust = 0.5, face = "bold"),
      legend.position = "right"
    )

  return(p)
}

# Create heatmap for hybrid design
p_hybrid <- plot_power_heatmap_single(summary_results, "hybrid")
print(p_hybrid)
```

### Multi-Design Comparison

```{r multi-heatmap}
# ============================================================================
# MULTI-DESIGN HEATMAP
# ============================================================================

plot_power_heatmap_multi <- function(data) {
  design_labels <- c(
    "ol_bdc" = "OL+BDC (Design 2)",
    "hybrid" = "Hybrid/N-of-1 (Design 4)"
  )

  plot_data <- data %>%
    mutate(
      design = factor(
        design_labels[as.character(design)],
        levels = design_labels
      ),
      carryover = factor(carryover),
      biomarker_moderation = factor(
        biomarker_moderation,
        levels = rev(sort(unique(biomarker_moderation)))
      )
    )

  p <- ggplot(
    plot_data,
    aes(x = carryover, y = biomarker_moderation, fill = power)
  ) +
    geom_tile(color = "white", linewidth = 0.5) +
    geom_text(
      aes(label = sprintf("%.0f%%", power * 100)),
      color = "black",
      size = 3.5,
      fontface = "bold"
    ) +
    scale_fill_gradient2(
      name = "Power",
      low = "#d73027",
      mid = "#fee08b",
      high = "#1a9850",
      midpoint = 0.5,
      limits = c(0, 1),
      labels = scales::percent
    ) +
    scale_x_discrete(expand = c(0, 0)) +
    scale_y_discrete(expand = c(0, 0)) +
    facet_wrap(~design, ncol = 2) +
    labs(
      title = "Statistical Power - Clustered Designs",
      subtitle = "N=70 participants, 100 iterations per condition",
      x = "Carryover (proportion retained)",
      y = "Biomarker Moderation (effect size)"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      panel.grid = element_blank(),
      axis.ticks = element_blank(),
      plot.title = element_text(hjust = 0.5, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5),
      legend.position = "right",
      strip.text = element_text(face = "bold", size = 12),
      strip.background = element_rect(fill = "grey90", color = NA)
    )

  return(p)
}

# Create multi-panel heatmap
p_multi <- plot_power_heatmap_multi(summary_results)
print(p_multi)
```

## Part 4: Effect Size Distributions

```{r effect-distributions}
# ============================================================================
# EFFECT SIZE DISTRIBUTIONS
# ============================================================================

# Filter to power conditions only (exclude Type I error conditions)
power_conditions <- results %>%
  filter(biomarker_moderation > 0)

# Plot effect size distributions by design
p_effect <- ggplot(power_conditions, aes(x = effect_size, fill = design)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~biomarker_moderation, labeller = label_both) +
  scale_fill_viridis_d(
    name = "Design",
    labels = c("Hybrid", "OL+BDC")
  ) +
  labs(
    title = "Treatment-Biomarker Interaction Effect Size Distributions",
    subtitle = "By biomarker moderation strength",
    x = "Effect Size (regression coefficient)",
    y = "Density"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )

print(p_effect)
```

### Effect Size vs Power

```{r effect-vs-power}
# Effect size vs power relationship
p_effect_power <- summary_results %>%
  filter(biomarker_moderation > 0) %>%
  ggplot(aes(x = mean_effect, y = power, color = factor(carryover))) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed") +
  facet_wrap(~design, labeller = label_both) +
  scale_color_viridis_d(name = "Carryover") +
  labs(
    title = "Effect Size vs Statistical Power",
    x = "Mean Effect Size",
    y = "Power"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )

print(p_effect_power)
```

## Part 5: Carryover Impact Analysis

```{r carryover-impact}
# ============================================================================
# CARRYOVER IMPACT ANALYSIS
# ============================================================================

# Calculate power change with carryover
carryover_impact <- summary_results %>%
  filter(biomarker_moderation > 0) %>%
  select(design, biomarker_moderation, carryover, power) %>%
  pivot_wider(
    names_from = carryover,
    values_from = power,
    names_prefix = "co_"
  ) %>%
  mutate(
    power_loss_50 = co_0 - co_0.5,
    power_loss_100 = co_0 - co_1,
    pct_loss_50 = (co_0 - co_0.5) / co_0 * 100,
    pct_loss_100 = (co_0 - co_1) / co_0 * 100
  )

cat("\n")
cat("=" %+% strrep("=", 59) %+% "\n")
cat("CARRYOVER IMPACT ON POWER\n")
cat("=" %+% strrep("=", 59) %+% "\n\n")

cat("Power loss due to carryover (relative to no carryover):\n\n")
print(carryover_impact)

# Visualize carryover impact
p_carryover <- summary_results %>%
  filter(biomarker_moderation > 0) %>%
  ggplot(aes(x = factor(carryover), y = power, group = biomarker_moderation,
             color = factor(biomarker_moderation))) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  facet_wrap(~design, labeller = label_both) +
  scale_color_viridis_d(name = "Biomarker\nModeration") +
  labs(
    title = "Power Degradation with Carryover",
    x = "Carryover (proportion retained)",
    y = "Statistical Power"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )

print(p_carryover)
```

## Part 6: Save Visualizations

```{r save-plots}
# ============================================================================
# SAVE VISUALIZATIONS
# ============================================================================

output_dir <- "../analysis/output"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

timestamp <- format(Sys.time(), "%Y%m%d_%H%M%S")

# Save multi-panel heatmap
ggsave(
  file.path(output_dir, sprintf("vignette_power_heatmap_%s.pdf", timestamp)),
  p_multi,
  width = 10,
  height = 7
)

# Save effect distribution
ggsave(
  file.path(output_dir, sprintf("vignette_effect_dist_%s.pdf", timestamp)),
  p_effect,
  width = 10,
  height = 6
)

# Save carryover impact
ggsave(
  file.path(output_dir, sprintf("vignette_carryover_impact_%s.pdf", timestamp)),
  p_carryover,
  width = 10,
  height = 5
)

cat("\n")
cat("=" %+% strrep("=", 59) %+% "\n")
cat("VISUALIZATION COMPLETE\n")
cat("=" %+% strrep("=", 59) %+% "\n\n")

cat("Saved to:", output_dir, "\n")
cat(sprintf("- vignette_power_heatmap_%s.pdf\n", timestamp))
cat(sprintf("- vignette_effect_dist_%s.pdf\n", timestamp))
cat(sprintf("- vignette_carryover_impact_%s.pdf\n", timestamp))
```

## Part 7: Detailed Results Table

```{r results-table}
# ============================================================================
# DETAILED RESULTS TABLE
# ============================================================================

cat("\n")
cat("=" %+% strrep("=", 59) %+% "\n")
cat("FULL RESULTS TABLE\n")
cat("=" %+% strrep("=", 59) %+% "\n\n")

formatted_results <- summary_results %>%
  mutate(
    power_pct = sprintf("%.1f%%", power * 100),
    effect_summary = sprintf("%.3f (%.3f)", mean_effect, sd_effect)
  ) %>%
  select(
    design, biomarker_moderation, carryover,
    power_pct, effect_summary, n
  ) %>%
  arrange(design, biomarker_moderation, carryover)

print(formatted_results, n = 50)
```

## Session Info

```{r session-info}
sessionInfo()
```

## References

Hendrickson RC, Thomas RG, Schork NJ, Raskind MA (2020). Optimizing Aggregated
N-Of-1 Trial Designs for Predictive Biomarker Validation: Statistical Methods
and Theoretical Findings. *Frontiers in Digital Health*, 2:13.
doi: 10.3389/fdgth.2020.00013
