---
title: "Vignette 3: Apply to Actual Clinical Trial Data"
author: "Rebecca C. Hendrickson, Ronald G. Thomas"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Vignette 3: Apply to Actual Clinical Trial Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  eval = TRUE
)
```

## Overview

This vignette demonstrates how to apply the pmsimstats2025 analysis methodology
to actual clinical trial data. The workflow covers:

1. Preparing clinical trial data for analysis
2. Defining the trial design structure
3. Fitting mixed-effects models with biomarker moderation
4. Generating personalized treatment response predictions
5. Visualizing individual patient outcomes

## Setup

```{r load-packages, message=FALSE}
rm(list = ls())
suppressPackageStartupMessages({
  library(tidyverse)
  library(nlme)
  library(MASS)
  library(conflicted)
})

conflicts_prefer(dplyr::select)
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)
```

## Part 1: Define Trial Design

### Trial Structure

Before analyzing clinical data, define the measurement schedule and treatment
assignment structure used in the trial.

```{r define-trial-design}
# ============================================================================
# TRIAL DESIGN SPECIFICATION
# ============================================================================

# Example: Hybrid N-of-1 design with 8 measurement points
measurement_weeks <- c(4, 8, 9, 10, 11, 12, 16, 20)

# Define treatment paths (4-path randomization)
# Path determines treatment assignment at each timepoint
treatment_paths <- tibble(
  path = 1:4,
  description = c(
    "Path 1: On -> Off -> On -> Off",
    "Path 2: On -> Off -> Off -> On",
    "Path 3: On -> Off (early) -> On -> Off",
    "Path 4: On -> Off (early) -> Off -> On"
  )
)

cat("Trial Design: Hybrid N-of-1\n")
cat("Measurement weeks:", paste(measurement_weeks, collapse = ", "), "\n")
cat("Number of paths:", nrow(treatment_paths), "\n")
```

### Create Design Matrix

```{r design-matrix}
# ============================================================================
# CREATE DESIGN MATRIX
# ============================================================================

create_trial_design_matrix <- function(measurement_weeks, n_paths = 4) {
  # Create full factorial of weeks and paths
  design <- expand_grid(
    path = 1:n_paths,
    week = measurement_weeks
  ) |>
    mutate(
      # Hybrid design treatment assignment rules
      treatment = case_when(
        week %in% c(4, 8) ~ 1,
        week == 9 ~ 1,
        week == 10 & path %in% c(1, 2) ~ 1,
        week == 10 & path %in% c(3, 4) ~ 0,
        week %in% c(11, 12) ~ 0,
        week == 16 & path %in% c(1, 3) ~ 1,
        week == 16 & path %in% c(2, 4) ~ 0,
        week == 20 & path %in% c(1, 3) ~ 0,
        week == 20 & path %in% c(2, 4) ~ 1,
        TRUE ~ NA_real_
      ),
      # Expectancy (blinding) status
      expectancy = if_else(week %in% c(4, 8), 1.0, 0.5),
      # Timepoint index
      timepoint = match(week, measurement_weeks)
    )

  return(design)
}

design_matrix <- create_trial_design_matrix(measurement_weeks)

cat("\nDesign Matrix Preview:\n")
print(head(design_matrix, 16))
```

## Part 2: Prepare Clinical Data

### Required Data Format

Clinical trial data must be in long format with the following columns:

| Column | Description |
|:-------|:------------|
| participant_id | Unique participant identifier |
| biomarker | Baseline biomarker value |
| baseline | Baseline symptom severity |
| path | Randomization path (1-4) |
| week | Measurement week |
| response | Symptom score at measurement |

```{r load-clinical-data}
# ============================================================================
# LOAD CLINICAL DATA
# ============================================================================

# Example: Load clinical trial data
# In practice, replace with your actual data file
# clinical_data <- read_csv("path/to/clinical_data.csv")

# For demonstration, generate synthetic clinical data
set.seed(42)
n_participants <- 50

# Generate participant-level data
participant_info <- tibble(
  participant_id = 1:n_participants,
  biomarker = rnorm(n_participants, mean = 5, sd = 2),
  baseline = rnorm(n_participants, mean = 10, sd = 2),
  path = sample(1:4, n_participants, replace = TRUE)
)

# Merge with design matrix
clinical_data <- participant_info |>
  left_join(design_matrix, by = "path", relationship = "many-to-many") |>
  mutate(
    # Simulate responses based on model
    bm_centered = (biomarker - mean(biomarker)) / sd(biomarker),
    BR_rate = 0.5 * (1 + 0.3 * bm_centered),
    weeks_on_drug = cumsum(treatment),
    response = baseline +
      weeks_on_drug * BR_rate * treatment +
      0.2 * week * expectancy +
      0.1 * week +
      rnorm(n(), 0, 1.5)
  ) |>
  select(participant_id, biomarker, baseline, path, week, response)

cat("Clinical data dimensions:", nrow(clinical_data), "rows x",
    ncol(clinical_data), "columns\n")
cat("\nFirst few observations:\n")
print(head(clinical_data))
```

### Data Quality Checks

```{r data-quality}
# ============================================================================
# DATA QUALITY CHECKS
# ============================================================================

cat("\n")
cat("=" %+% strrep("=", 59) %+% "\n")
cat("DATA QUALITY CHECKS\n")
cat("=" %+% strrep("=", 59) %+% "\n\n")

# Check for missing values
missing_summary <- clinical_data |>
  summarize(across(everything(), ~sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "column", values_to = "n_missing")

cat("Missing values:\n")
print(missing_summary)

# Check participant counts
cat("\nParticipants per path:\n")
participant_paths <- clinical_data |>
  distinct(participant_id, path) |>
  count(path)
print(participant_paths)

# Check timepoints per participant
cat("\nTimepoints per participant (should be 8):\n")
timepoints_check <- clinical_data |>
  count(participant_id) |>
  count(n, name = "n_participants")
print(timepoints_check)
```

## Part 3: Fit Mixed-Effects Model

### Model Specification

The analysis uses a linear mixed-effects model with:

- Fixed effects: treatment, biomarker, treatment × biomarker interaction, week
- Random effects: participant-specific intercept
- Correlation structure: CAR(1) for repeated measures

```{r fit-model}
# ============================================================================
# FIT MIXED-EFFECTS MODEL
# ============================================================================

# Prepare data for modeling
model_data <- clinical_data |>
  left_join(design_matrix, by = c("path", "week")) |>
  mutate(
    bm_centered = biomarker - mean(biomarker),
    participant_id = factor(participant_id)
  )

# Fit LME model with biomarker × treatment interaction
model_fit <- lme(
  response ~ treatment * bm_centered + week,
  random = ~ 1 | participant_id,
  correlation = corCAR1(form = ~ week | participant_id),
  data = model_data,
  control = lmeControl(opt = "optim")
)

cat("\n")
cat("=" %+% strrep("=", 59) %+% "\n")
cat("MODEL RESULTS\n")
cat("=" %+% strrep("=", 59) %+% "\n\n")

# Model summary
summary(model_fit)
```

### Extract Coefficients

```{r extract-coefs}
# ============================================================================
# EXTRACT AND INTERPRET COEFFICIENTS
# ============================================================================

coef_table <- summary(model_fit)$tTable
coef_df <- as.data.frame(coef_table) |>
  rownames_to_column("term") |>
  as_tibble()

cat("\nFixed Effects:\n")
print(coef_df)

# Test for biomarker × treatment interaction
interaction_p <- coef_df |>
  filter(term == "treatment:bm_centered") |>
  pull(`p-value`)

cat("\n")
cat("=" %+% strrep("=", 59) %+% "\n")
cat("BIOMARKER MODERATION TEST\n")
cat("=" %+% strrep("=", 59) %+% "\n\n")

cat(sprintf("Interaction p-value: %.4f\n", interaction_p))
if (interaction_p < 0.05) {
  cat("Result: SIGNIFICANT biomarker moderation of treatment effect\n")
  cat("Interpretation: Treatment response varies by biomarker level\n")
} else {
  cat("Result: No significant biomarker moderation detected\n")
  cat("Interpretation: Treatment effect is consistent across biomarker levels\n")
}
```

## Part 4: Generate Predictions

### Predicted Treatment Response

```{r generate-predictions}
# ============================================================================
# GENERATE PREDICTIONS
# ============================================================================

# Create prediction grid over biomarker range
biomarker_range <- seq(
  min(clinical_data$biomarker),
  max(clinical_data$biomarker),
  length.out = 50
)

# Predictions for treatment ON vs OFF
prediction_grid <- expand_grid(
  bm_centered = biomarker_range - mean(clinical_data$biomarker),
  treatment = c(0, 1),
  week = max(measurement_weeks)
) |>
  mutate(
    biomarker = bm_centered + mean(clinical_data$biomarker)
  )

# Get population-level predictions (fixed effects only)
prediction_grid$predicted <- predict(
  model_fit,
  newdata = prediction_grid,
  level = 0  # Population level (no random effects)
)

# Calculate treatment effect as function of biomarker
treatment_effect_by_bm <- prediction_grid |>
  select(biomarker, treatment, predicted) |>
  pivot_wider(
    names_from = treatment,
    values_from = predicted,
    names_prefix = "trt_"
  ) |>
  mutate(
    treatment_effect = trt_1 - trt_0
  )

cat("\nPredicted Treatment Effect by Biomarker Level:\n")
print(head(treatment_effect_by_bm, 10))
```

## Part 5: Visualize Results

### Treatment Effect by Biomarker

```{r plot-treatment-effect}
# ============================================================================
# VISUALIZATION: TREATMENT EFFECT BY BIOMARKER
# ============================================================================

p_treatment_effect <- ggplot(
  treatment_effect_by_bm,
  aes(x = biomarker, y = treatment_effect)
) +
  geom_line(linewidth = 1.2, color = "#1a9850") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_rug(
    data = clinical_data |> distinct(participant_id, biomarker),
    aes(x = biomarker),
    inherit.aes = FALSE,
    sides = "b",
    alpha = 0.5
  ) +
  labs(
    title = "Predicted Treatment Effect by Biomarker Level",
    subtitle = "Positive values indicate symptom improvement with treatment",
    x = "Biomarker Value",
    y = "Treatment Effect\n(change in symptom score)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(color = "grey40")
  )

print(p_treatment_effect)
```

### Individual Patient Trajectories

```{r plot-trajectories}
# ============================================================================
# VISUALIZATION: INDIVIDUAL TRAJECTORIES
# ============================================================================

# Select subset of participants for visualization
sample_participants <- sample(unique(clinical_data$participant_id), 12)

# Add predictions to full model_data first, then filter
model_data$predicted <- predict(model_fit)

trajectory_data <- model_data |>
  filter(participant_id %in% sample_participants) |>
  mutate(
    treatment_label = if_else(treatment == 1, "On Drug", "Off Drug")
  )

p_trajectories <- ggplot(
  trajectory_data,
  aes(x = week, y = response, group = participant_id)
) +
  geom_line(alpha = 0.5) +
  geom_point(aes(color = treatment_label), size = 2) +
  geom_line(aes(y = predicted), linetype = "dashed", alpha = 0.7) +
  facet_wrap(~participant_id, ncol = 4) +
  scale_color_manual(
    values = c("On Drug" = "#1a9850", "Off Drug" = "#d73027"),
    name = "Treatment"
  ) +
  labs(
    title = "Individual Patient Trajectories",
    subtitle = "Solid = observed, Dashed = model predicted",
    x = "Week",
    y = "Symptom Score"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom",
    strip.text = element_text(size = 8)
  )

print(p_trajectories)
```

### Personalized Response Prediction

```{r personalized-prediction}
# ============================================================================
# PERSONALIZED PREDICTION FOR NEW PATIENT
# ============================================================================

predict_patient_response <- function(model, biomarker_value, baseline_value,
                                     treatment_weeks = c(4, 8, 9, 10)) {
  # Prepare prediction data
  mean_biomarker <- mean(clinical_data$biomarker)

  pred_data <- tibble(
    week = measurement_weeks,
    bm_centered = biomarker_value - mean_biomarker,
    treatment = if_else(week %in% treatment_weeks, 1, 0)
  )

  # Generate predictions
  pred_data$predicted <- predict(model, newdata = pred_data, level = 0)

  # Adjust for baseline
  baseline_adjustment <- baseline_value - pred_data$predicted[1]
  pred_data$predicted <- pred_data$predicted + baseline_adjustment

  return(pred_data)
}

# Example: Predict for a new patient
new_patient <- predict_patient_response(
  model = model_fit,
  biomarker_value = 7.5,
  baseline_value = 12.0
)

cat("\n")
cat("=" %+% strrep("=", 59) %+% "\n")
cat("PERSONALIZED PREDICTION EXAMPLE\n")
cat("=" %+% strrep("=", 59) %+% "\n\n")

cat("New patient profile:\n")
cat("  Biomarker: 7.5\n")
cat("  Baseline symptom: 12.0\n\n")

cat("Predicted trajectory:\n")
print(new_patient)
```

## Part 6: Model Diagnostics

```{r diagnostics}
# ============================================================================
# MODEL DIAGNOSTICS
# ============================================================================

# Residual plot
residuals_df <- tibble(
  fitted = fitted(model_fit),
  residuals = residuals(model_fit),
  participant_id = model_data$participant_id
)

p_resid <- ggplot(residuals_df, aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  geom_smooth(method = "loess", se = FALSE, color = "blue") +
  labs(
    title = "Residuals vs Fitted Values",
    x = "Fitted Values",
    y = "Residuals"
  ) +
  theme_minimal(base_size = 11)

print(p_resid)

# QQ plot
p_qq <- ggplot(residuals_df, aes(sample = residuals)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(
    title = "Normal Q-Q Plot of Residuals",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  ) +
  theme_minimal(base_size = 11)

print(p_qq)

# Autocorrelation check
cat("\nEstimated AR(1) correlation:\n")
cat(sprintf("  rho = %.3f\n", coef(model_fit$modelStruct$corStruct, unconstrained = FALSE)))
```

## Part 7: Export Results

```{r export-results}
# ============================================================================
# EXPORT RESULTS
# ============================================================================

output_dir <- "../analysis/output"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

# Save model object
saveRDS(model_fit, file.path(output_dir, "clinical_model_fit.rds"))

# Save coefficient table
write_csv(coef_df, file.path(output_dir, "clinical_model_coefficients.csv"))

# Save predictions
write_csv(treatment_effect_by_bm, file.path(output_dir, "treatment_effect_predictions.csv"))

cat("\n")
cat("=" %+% strrep("=", 59) %+% "\n")
cat("EXPORT COMPLETE\n")
cat("=" %+% strrep("=", 59) %+% "\n\n")

cat("Saved to:", output_dir, "\n")
cat("  - clinical_model_fit.rds\n")
cat("  - clinical_model_coefficients.csv\n")
cat("  - treatment_effect_predictions.csv\n")
```

## Session Info

```{r session-info}
sessionInfo()
```

## References

Hendrickson RC, Thomas RG, Schork NJ, Raskind MA (2020). Optimizing Aggregated
N-Of-1 Trial Designs for Predictive Biomarker Validation: Statistical Methods
and Theoretical Findings. *Frontiers in Digital Health*, 2:13.
doi: 10.3389/fdgth.2020.00013
