---
title: "Quickstart Guide: pmsimstats2025"
author: "Rebecca C. Hendrickson, Ronald G. Thomas"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quickstart Guide: pmsimstats2025}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

**pmsimstats2025** is an N-of-1 clinical trial simulation study comparing
multiple trial designs with a focus on statistical power when biomarker-
treatment interactions exist and carryover effects are present.

The project implements methodology from **Hendrickson et al. (2020)** with
extensions for explicit biomarker moderation and carryover effect modeling.

## Installation

```{r install}
# Clone the repository
# git clone https://github.com/rgt47/pmsimstats2025.git
# cd pmsimstats2025

# Restore R environment
renv::restore()
```

## Quick Start

### Run a Simulation

The simulation is organized into two main scripts based on measurement
schedule type:

```{r run-simulation}
# Clustered measurement schedule designs (Hybrid, OL+BDC)
source("analysis/scripts/simulation_clustered.R")

# Evenly-spaced designs (OL, Crossover, Parallel)
source("analysis/scripts/simulation_evenly_spaced.R")
```

### Generate Visualizations

```{r visualize}
# Power heatmaps from simulation results
source("analysis/scripts/visualize_hendrickson_style.R")
source("analysis/scripts/visualize_power_heatmaps_4panel.R")
```

## Trial Designs

The project supports five trial designs:

| Design | Points | Structure |
|:-------|:-------|:----------|
| Hybrid | 8 | Dense cluster at treatment transition |
| OL+BDC | 8 | Dense cluster at biomarker discontinuation |
| Open-Label | 4 | Evenly spaced (6-week gaps) |
| Crossover | 4 | Evenly spaced (6-week gaps) |
| Parallel | 4 | Evenly spaced (6-week gaps) |

### Measurement Schedule Rationale

Clustered schedules enable more measurement points (8) within a 20-week trial
by grouping observations around key transition times. Evenly-spaced designs
are limited to 4 points to maintain positive definite covariance matrices.

## Core Functions

### Parameter Grid Validation

Before running simulations, validate parameter combinations:
```{r validate}
source("analysis/scripts/pm_functions.R")

# Define parameter grid
param_grid <- expand.grid(
  biomarker_moderation = c(0.2, 0.4),
  carryover_t1half = c(0, 1.0, 2.0),
  n_participants = 70
)

# Validate all combinations
validation <- validate_parameter_grid(param_grid)
report_parameter_validation(validation)
```

### Build Sigma Matrix

```{r sigma}
# Construct covariance matrix for a trial design
sigma <- build_sigma_matrix(
  n_timepoints = 8,
  c.br = 0.75,
  c.tr = 0.75,
  c.er = 0.75
)
```

### Generate Trial Data

```{r generate}
# Generate data for one participant
trial_data <- generate_data(
  sigma = sigma,
  n_timepoints = 8,
  treatment_schedule = c(0, 0, 1, 1, 1, 1, 0, 0)
)
```

## Carryover Modeling

The simulation compares two analysis scenarios using `nlme::lme` with
continuous-time AR(1) correlation (`corCAR1`) to properly match the DGP:

### With Explicit Carryover (effective_drug_weeks)

```{r model-with}
library(nlme)

fit_with <- lme(
  response ~ effective_drug_weeks * bm_centered + week,
  random = ~ 1 | participant_id,
  correlation = corCAR1(form = ~ week | participant_id),
  data = trial_data
)
```

### Without Carryover Term

```{r model-without}
fit_without <- lme(
  response ~ treatment * bm_centered + week,
  random = ~ 1 | participant_id,
  correlation = corCAR1(form = ~ week | participant_id),
  data = trial_data
)
```

### Effective Drug Weeks Computation

Effective drug weeks incorporates exponential decay carryover:

```{r carryover}
# On treatment: cumulative weeks on drug
# Off treatment: weeks_at_discontinuation * (0.5)^(time_since_discont / t_half)
effective_drug_weeks <- weeks_on_drug * (0.5)^(tsd / t_half)
```

Where:

- `tsd` is time since discontinuation (weeks)
- `t_half` is the carryover half-life in weeks (0, 0.5, 1.0, or 2.0)

## Output Files

### Simulation Results

Results are saved as RData files:

- `analysis/output/simulation_clustered_results.RData`
- `analysis/output/simulation_evenly_spaced_results.RData`

Each file contains:

- `simulation_results` - One row per iteration with p-values and effect sizes
- `simulation_summary` - Aggregated power by condition
- `param_grid` - Parameter specifications

### Visualizations

Power heatmaps are saved as PDF files:

- `analysis/output/power_heatmap_clustered.pdf`
- `analysis/output/power_heatmap_evenly_spaced.pdf`
- `analysis/output/power_heatmap_4panel.pdf`

## Docker Workflow

For reproducibility, run simulations in Docker:

```bash
# Start container with all dependencies
make r

# Inside container, run simulations
Rscript analysis/scripts/simulation_clustered.R
```

## Modifying Parameters

### Change Biomarker Correlation Grid

Edit the parameter grid in simulation files:

```{r param-grid}
param_grid <- expand_grid(
  biomarker_moderation = c(0.2, 0.4, 0.6),
  carryover_t1half = c(0, 1.0, 2.0),
  n_participants = 70
)
```

### Adjust Iteration Count

Change `n_iterations` in simulation files:

```{r iterations}
n_iterations <- 100  # More stable estimates (slower)
```

## Troubleshooting

### Non-Positive Definite Matrix

If you see "Non-positive definite matrix rejected", this is normal behavior
for invalid parameter combinations. Check:

1. Biomarker correlation values (>0.4 may cause issues)
2. Measurement schedule spacing
3. Condition number output

### Convergence Warnings

Occasional mixed model convergence warnings are expected. Monitor the `error`
column in results if >20% of iterations fail.

## References

Hendrickson, E., et al. (2020). N-of-1 trials with multiple randomization
structures for individualized treatment. *Statistics in Medicine*,
39(25), 3581-3599.

## Next Steps

- `analysis/scripts/README.md` - Detailed workflow guide
- `docs/simulation_white_paper.md` - Methodology documentation
- `docs/sigma_matrix_derivation.tex` - Mathematical foundations
